---
title: "Sparse-constrained GLM"
subtitle: "The problem we want to solve"
---

## A familiar starting point: bodyfat

Suppose we want to predict body-fat percentage from physical measurements. The classical dataset (Penrose et al. 1985) records 14 measurements on 252 men:

```{r}
#| label: bodyfat-table
#| echo: false
df <- data.frame(
  Age      = c(23, 22, 22, 26, 24),
  Weight   = c(70.1, 78.8, 70.0, 83.5, 80.4),
  Height   = c(172.1, 183.5, 168.3, 183.5, 180.3),
  Neck     = c(36.2, 38.5, 34.0, 37.4, 39.0),
  Chest    = c(93.1, 93.6, 95.8, 101.8, 97.3),
  Abdomen  = c(85.2, 83.0, 87.9, 86.4, 100.0),
  Hip      = c(94.5, 98.7, 99.2, 101.2, 101.9),
  Thigh    = c(59.0, 58.7, 59.6, 60.1, 63.2),
  Knee     = c(37.3, 37.3, 38.9, 37.3, 42.2),
  Ankle    = c(21.9, 23.4, 24.0, 22.8, 24.0),
  Biceps   = c(32.0, 30.5, 28.8, 32.4, 32.2),
  Forearm  = c(27.4, 28.9, 25.2, 29.4, 27.7),
  Wrist    = c(17.1, 18.2, 16.6, 18.2, 17.7),
  Bodyfat  = c(12.3, 6.1, 25.3, 10.4, 28.7)
)
knitr::kable(df, caption = "Bodyfat data — first 5 rows. Response **Bodyfat** (%) in last column.")
```

Which **handful** of these measurements is enough to predict body fat well? That is best subset selection.

## Generalising to GLMs

The same question arises whenever the response is well modelled by a generalised linear model (GLM):

| Response type | Family | Examples |
|---|---|---|
| Continuous | Gaussian | bodyfat, gene-expression intensity |
| Binary | Bernoulli (logistic) | disease yes/no, GWAS case/control |
| Multinomial | softmax (multinomial logistic) | cancer subtypes |
| Count | Poisson | event counts |

In every case, fitting the model amounts to maximising a log-likelihood:

$$
\ell(\beta_0, \boldsymbol{\beta})
\;=\; \sum_{i=1}^n \log p(y_i \mid x_i; \beta_0, \boldsymbol{\beta}).
$$

For Gaussian, this reduces to least squares; for Bernoulli, to logistic regression; and so on. The unifying framework lets one method handle them all — provided it can deal with the **sparsity constraint** we are about to add.

## The optimisation we are trying to solve

For a prescribed model size $k$, find the $k$ predictors that best fit the data:

$$
\boxed{\;\;
\min_{\beta_0 \in \mathbb{R},\; \boldsymbol{\beta} \in \mathbb{R}^p}\;
\;-\tfrac{1}{n}\,\ell(\beta_0, \boldsymbol{\beta})
\quad
\text{subject to}\quad
\|\boldsymbol{\beta}\|_0 \le k.
\;\;}
$$

Two things to notice:

- The objective is the **negative GLM log-likelihood**.
- The constraint is the **$\ell_0$-norm cardinality**: count the non-zero entries, keep at most $k$.

The user supplies $k$. Output: which $k$ features, and their fitted coefficients.

This is the problem COMBSS solves.

## Why it is hard

Counting feasible subsets:

```{r}
#| label: combinatorial-counts
#| echo: false
counts <- data.frame(
  Setting = c("Bodyfat (p=13, k=5)",
              "Khan SRBCT (p=2308, k=12)",
              "Rice GWAS (p=158210, k=10)"),
  `Candidate subsets` = c(
    format(choose(13, 5), big.mark = ","),
    format(signif(choose(2308, 12), 3), big.mark = ","),
    format(signif(choose(158210, 10), 3), big.mark = ",")
  ),
  check.names = FALSE
)
knitr::kable(counts)
```

For $p = 13$ enumeration is fine. For $p = 2308$ it is **$10^{33}$ subsets**; for $p \approx 1.6 \times 10^5$ it is **$10^{45}$**. The combinatorial blowup makes exhaustive search hopeless within the lifetime of the universe.

In fact the problem is **NP-hard** (Natarajan 1995). No polynomial-time algorithm is expected for the worst case.

## Where this matters

The setting $p \gg n$ with a sparse truth is the bread and butter of modern applied statistics:

- **Genomics** — GWAS with $10^5$ SNPs, gene-expression panels with $10^3$–$10^4$ probes.
- **Survey data** — patient registries with hundreds of recorded variables and a few hundred outcomes.
- **Sensor data** — many channels, few labelled trials.
- **Financial markets** — predicting asset returns or default risk from hundreds of candidate factors (macroeconomic indicators, fundamentals, sentiment, technical signals); a small interpretable factor model is preferred over a dense black-box.

Every one of these calls for a sparsity-constrained GLM, not just a regularised one.

## The plan from here

How do existing methods cope with the NP-hard constraint? Two main strategies:

- **Lasso** ([next page](02-lasso.qmd)) — relax $\|\boldsymbol{\beta}\|_0$ to the convex $\|\boldsymbol{\beta}\|_1$. Fast and popular, but biased and indexed by $\lambda$, not $k$.
- **MIO** ([page after](03-mio.qmd)) — solve the discrete problem directly with mixed-integer optimisation. Exact, but does not scale.

COMBSS sits between the two: continuous like the lasso, but on the *support indicator* rather than on $\boldsymbol{\beta}$ itself — and explicitly $k$-indexed.

---

[Next → Lasso](02-lasso.qmd)
